Simple and Compositional Reification of Monadic Embedded Languages

### 2013-09-30 src2/bb/partINew.tex
another approach

\section{Introduction}
\studname{} is a computer science student working in a project involving programming 
robots in a low-level imperative language. However, \studname{} has a budding 
interest in functional programming using Haskell and has read ``The Haskell 
school of expression'' \cite{HUDAK}. He gets the idea to implement 
a language for robot control embedded in Haskell. \studname{} realises that the 
capabilities of the robot hardware are very similar to those of the 
robots that Hudak evaluates graphically in a grid world. There is however 
a very important difference; the embedded language needs to be compiled into 
some form that is understood by the robot. 
%\studname{} in a computer science student with a budding interest in functional 
%programming using Haskell. \studname{} is developing  
%an embedded domain specific language (EDSL) for robot control. Inspired 
%by the book ``the Haskell school of expression'' \cite{HUDAK} but instead of 
%evaluating his robot EDSL, \studname{} wants to compile the language to a form 
%he can run on actual robot hardware. 
%with 
%different motivations, \studname{} wants compile his EDSL for actual physical robots, 
%he sets out.

Guided by the capabilities of the target robot, \studname{} designs an API for 
robot programming. The robot can perform operations 
such as {\em move} that steps the robot forward and {\em turn} left and right. 
The robot also has the capability to execute program loops and conditionals and 
it has a forward facing {\tt sensor} with which it can query the world.

%The robot also has a forward facing {\tt sensor} with which it can query the 
%world and also the capabilities to execute program loops and conditionals. 
\pagebreak
The API that \studname{} designs is simple:

%\begin{figure} 
\begin{small}
\begin{verbatim} 
move      :: Program () 
turnLeft  :: Program () 
turnRight :: Program ()
sensor    :: Program Bool 
cond      :: Program Bool 
             -> Program () 
             -> Program () 
             -> Program () 
while     :: Program Bool 
             -> Program () 
             -> Program () 
\end{verbatim}
\end{small} 
%\label{fig:interface} 
%\caption{Proposed set of basic robot operations}
%\end{figure}

@@

To be able to create a first-order representation of a computation
described using monads, \studname{} needs to solve the problem of {\em
  reifying} monads. This is the story of how \studname{} came up with
a particularly simple solution to this problem.

%\emph{*** Josef, Think about writing here}

%\begin{minipage}[]{\linewidth}
%\vspace{5mm}

%\fbox{%
%\parbox{0.8\linewidth}{What \studname{} does not know at this time is that he has stumbled over the 
%problem of {\em Monad reification}. Monad reification is the observing of the shape of the computation and this is necessary in order to compile to some low level representation.}
%}
%\end{minipage}

\subsection{Example programs}

\studname{} is quite happy with his language design and decides to
write some programs to try it out.
%experiments 
%with what some of these robot controlling programs could look like once his 
%EDSL is finished.

The first program is called {\tt sMove}, which implements a safe move 
operation. This operation can be executed by the robot even though it is facing 
an obstacle, without risk of harming robot or obstacle. 
%.It feels like a good idea to \studname{} to check the sensor before trying to 
%move the robot forward. Otherwise something in either the world or the 
%robot might suffer damage.

\begin{small} 
\begin{verbatim}
sMove :: Program () 
sMove = cond sensor turnRight move 
\end{verbatim}
\end{small}

The second program moves the robot forward until it stands directly 
in front of an obstacle such as a wall.

@@

% Since an abstract syntax representation of the robot programs is desired,
Since the language is going to be compiled, \studname{} realises that the 
booleans in his language cannot be regular Haskell booleans. The booleans 
needs to be replaced by boolean typed expressions.

\pagebreak
\begin{small} 
\begin{verbatim}
type Name = String

data BoolE = Lit Bool
           | Var Name
           | (:||:) BoolE BoolE
           | (:&&:)  BoolE BoolE
           | Not BoolE  
\end{verbatim} 
\end{small}

Following this, the  operations that should go into the {\tt Program} data type 
feel straightforward.

\begin{small} 
\begin{verbatim}
data Program a where 
  Move      :: Program () 
  TurnLeft  :: Program () 
  TurnRight :: Program () 
  Sensor    :: Program BoolE
  Cond      :: Program BoolE 
               -> Program () 
               -> Program () 
  While     :: Program BoolE 
               -> Program () 
               -> Program () 
\end{verbatim} 
\end{small}

\studname{} continues by naively adding constructors for {\tt Return} and 
{\tt Bind} to the {\tt Program} data type.

@@

\subsection{Problem statement} 
\emph{Meanwhile in Professor \docname{}'s Office}\newline %\newline

@@

\noindent\colorbox{light-gray}{
\begin{minipage}[]{0.9\linewidth}
\noindent 
Dear Professor \docname{}
\newline \newline
\noindent I am CS student working in a robot control project. Usually, we 
program our robots in C but I have developed an interest in Haskell 
programming and thought it'd be natural to try implementing an EDSL. 
I know you teach an FP course and thought I would ask for your input. 
Attached to this mail is a file containing an outline of the data types I 
want to use. Does this look sensible to you? \newline \newline

@@

Prof. \docname{} opens the attachment and takes a look at the data type. He 
particularly notices the constructors {\tt Return} and {\tt Bind} and the 
{\tt Monad} instance. He shakes his head at \studname{}'s naivet\'e and writes an 
email back:

%\vspace{5mm}

\noindent\colorbox{light-gray}{
\begin{minipage}[]{0.9\linewidth}
\noindent 
Hello \studname{}
\newline \newline
\noindent I'm afraid your implementation of the monadic primitives can never 
work. The constructor {\tt Return} can take any arbitrary value that has 
nothing to do with the language you're designing. These values may be 
strings, binary trees or higher-order functions from zygomorphisms to 
histomorphisms. The same problem goes for {\tt Bind}; it has to be able 
to handle arbitrary values. It cannot possibly work! This is a very 
difficult problem and such a naive solution is bound to fail. 
\newline \newline
%% \noindent For hints towards a solution to this I suggest that you follow 
%% this \underline{link} \cite{GillBlog}. Another idea is that you refer 
%% to the article ``Generic monadic constructs for embedded languages'' 
%% \cite{Generic}.
%% \newline \newline

@@

% \section{Implementation continues}
% \emph{Back to \studname{}'s implementation efforts} \newline \newline

@@

When \docname{}'s email does arrive, \studname{} is confused. He feels he 
has already managed to compile the {\tt Program} data type into a 
representation closer to that which is executed by the robot hardware.

@@

\noindent\colorbox{light-gray}{
\begin{minipage}[]{0.9\linewidth}
\noindent 
Hello Prof. \docname{} 
\newline \newline
\noindent Thanks for your feedback. But I think it does work! Attached
to this email you find a file containing my attempt at compiling the
{\tt Program} data type into a representation closer to that which is
executed by our robots. \newline \newline

\noindent I have also attached two example programs that you can compile and 
then run in the graphical simulator (see figure~\ref{fig:programs}) \newline \newline
%\vspace{5mm}

%\includegraphics[width=.49\linewidth]{./spiral1}
%\includegraphics[width=.49\linewidth]{./spiral2}

%\vspace{5mm}

%\noindent And these two images are output from running the 
%{\tt followWall} (see figure~\ref{fig:followWall}) program. \newline

%\vspace{5mm}

%\includegraphics[width=.49\linewidth]{./wall1}
%\includegraphics[width=.49\linewidth]{./wall2}

%\vspace{5mm}

@@

\FloatBarrier

\begin{figure*}
\begin{tabular}{c|c}
\begin{minipage}[]{0.45\columnwidth}
\begin{small}
\begin{verbatim}
spiralIn :: Int -> Program () 
spiralIn 0 = return () 
spiralIn n = do
  replicateM_ 2 $ do
    replicateM_ n move
    turnLeft
  spiralIn (n-1) 
\end{verbatim}
\end{small}
\end{minipage}
\hspace{0.05\columnwidth} 
\begin{minipage}[]{0.45\columnwidth}
\includegraphics[width=\columnwidth]{./bb/spiral1}
\end{minipage}
&
\begin{minipage}[]{0.45\columnwidth}
\begin{small}
\begin{verbatim} 
followWall :: Program () 
followWall = 
  while (return true) $ 
    cond checkLeft sMove $ 
            do turnLeft
               move

checkLeft :: Program BoolE     
checkLeft = do
  TurnLeft
  s <- Sensor
  TurnRight 
  return s
\end{verbatim}
\end{small}
\end{minipage}
\hspace{0.05\columnwidth}
\begin{minipage}[]{0.45\columnwidth}
\includegraphics[width=\columnwidth]{./bb/wall2}
\end{minipage}

\end{tabular}
\caption{The {\tt spiralIn} and {\tt followWall} programs sent from \studname{}
to \docname{} in an email}
\label{fig:programs}
\end{figure*}

\section{Compilation of the monadic robot EDSL} 
\emph{\docname{} receives \studname{}'s latest email}\newline %\newline

\noindent \docname{} looks through the {\tt Compiler} module and finds a data type describing
a first-order representation of the robot language.

\begin{small} 
\begin{verbatim}
data Prg = PMove          
         | PTurnRight
         | PTurnLeft
         | PSensor Name  
         | PCond BoolE Prg Prg
         | PWhile Name Prg Prg 
         | PSeq Prg Prg
         | PSkip       
         | PAssign Name BoolE
\end{verbatim}
\end{small}

\begin{figure} 
\begin{small}
\begin{verbatim} 
newNameSupply :: NameSupply 
split2 :: NameSupply -> 
          (NameSupply,NameSupply) 
split3 :: NameSupply -> 
          (NameSupply,NameSupply,NameSupply)
supplyValue :: NameSupply -> Int 
\end{verbatim} 
\end{small}
\caption{The interface of the splittable name supply used in the implementation 
of the compile functions} 
\label{fig:namesupply}
\end{figure}

From the {\tt Prg} data type, \docname{} presumes that the {\tt PSeq}
constructor will be the result of compiling {\tt Bind} and that
{\tt PSensor} binds a variable which can be used in {\tt PAssign} and
{\tt PWhile}. The conclusion is that the representation looks
sensible, but he is very interested in seeing the {\tt Program a -> Prg}
transformation. The compile function assumes a splittable name
supply with the interface described in figure \ref{fig:namesupply}.

\pagebreak
\begin{small} 
\begin{verbatim}
runCompile :: Program a -> Prg
runCompile prg = snd $ compile s prg 
  where 
    s = newNameSupply

compile :: NameSupply -> Program a -> (a, Prg) 
compile s Move = ((),PMove)
compile s TurnRight = ((),PTurnRight)
compile s TurnLeft  = ((),PTurnLeft)
compile s Sensor    = (Var nom,PSensor nom)
  where
    v   = supplyValue s
    nom = "v" ++ show v 
compile s (Cond b p1 p2) = 
    ((),bp `PSeq` PCond b' p1' p2') 
  where
    (s1,s2,s3) = split3 s
    (b',bp)    = compile s1 b
    (a1,p1')   = compile s2 p1
    (a2,p2')   = compile s3 p2
compile s (While wp prg) = ((),PWhile nom nwp prg') 
  where
    (s1,s2,s3) = split3 s
    (b,wp')    = compile s1 wp
    (c,prg')   = compile s2 prg 
    nom        = "v" ++ (show $ supplyValue s3)
    nwp        = (wp' `PSeq` PAssign nom b) 
compile s (Return a) = (a,PSkip)
compile s (Bind pa f) = (b, prg1 `PSeq` prg2) 
  where
    (s1,s2)  = split2 s
    (a,prg1) = compile s1 pa
    (b,prg2) = compile s2 (f a) 
\end{verbatim}
\end{small}

%\begin{small} 
%\begin{verbatim}
%compile :: Program a -> Prg
%compile prg = snd $ compile' s prg 
%  where 
%    s = unsafePerformIO $ newEnumSupply
%    
%compile' :: Supply Int -> Program a -> (a, Prg) 
%compile' s Move = ((),PMove)
%compile' s TurnRight = ((),PTurnRight)
%compile' s TurnLeft  = ((),PTurnLeft)
%compile' s Sensor    = (Var nom,PSensor nom)
%  where
%    v = supplyValue s
%    nom = "v" ++ show v 
%compile' s (Cond b p1 p2) = ((), PCond b p1' p2') 
%  where
%    (s1,s2) = split2 s
%    (a1,p1') = compile' s1 p1
%    (a2,p2') = compile' s2 p2 
%compile' s (While wp prg) = ((),PWhile nom nwp prg') 
%  where
%    (s1,s2,s3) = split3 s
%    (b,wp') = compile' s1 wp
%    (c,prg') = compile' s2 prg 
%    nom  = "v" ++ (show $ supplyValue s3)
%    nwp = (wp' `PSeq` PAssign nom b) 
%compile' s (Return a) = (a,PSkip)
%compile' s (Bind pa f) = (b, prg1 `PSeq` prg2) 
%  where
%    (s1,s2) = split2 s
%    (a,prg1) = compile' s1 pa
%    (b,prg2) = compile' s2 (f a) 
%\end{verbatim}
%\end{small}

\docname{} studies the code in disbelief, saves the module and tries
it out on some examples. The code does indeed seem to work and
\docname{}'s disbelief is replaced with enthusiasm; this appears to be
a simple and convenient way to reify monads.
%What's more it does so in a 
%compositional way. The compilation of the {\tt Return} and {\tt Bind} cases 
%are completely independent of any other construct of the language.

@@

\section{Meeting in \docname{}'s office} %\section{Related Work} 
%\emph{Meeting in \docname{}'s office} \newline

@@

\speak{\docname{}} So, about this robot language! You gave me a bit of a 
surprise there. I was entirely sure that what you did was impossible.

@@

\speak{\docname{}} The problem is \direct{\docname{} approaches his whiteboard}
 when you try to reify the bind constructor of your representation.
\end{dialogue}

@@

\begin{dialogue}
\speak{\docname{}} You need to come up with an element of type {\tt a} to 
pass to the function.

@@

\speak{\docname{}} But it is! Your solution is that you are careful
about the return types of all operations in your language. They are
either of unit type or some type which can be guaranteed to be reified.
\direct{\docname{} scribbles on his whiteboard}

@@

\begin{dialogue}
\speak{\docname{}} This is different from how such constructs are
normally implemented. If I had implemented the language I would have
given {\tt Sensor} the type {\tt Program Bool} so that I could write
an evaluator of type {\tt Program a -> a}, or perhaps {\tt Program a -> M a}.
But then adding a monad to the language and reifying it becomes much harder.

\speak{\studname{}} Yes, now that you mention it, I have seen that style used
for defining DSLs.

\speak{\docname{}} Having constructors like {\tt Sensor} return {\tt
  BoolE} instead of {\tt Bool} is a crucial part of why your
method works. It allows the {\tt compile} function to be written in
such a way that it can generate the return values statically and pass
them as arguments to {\tt Bind}. In a sense you're doing evaluation at
the same time as compilation. There's just enough evaluation to
remove the monadic constructs. It's a very neat trick!

\speak{\studname{}} Thank you.

\speak{\docname{}} Another way to think about your trick is that you are
using a writer monad to evaluate your language and produce the first-order
syntax tree as a side effect. Then you could have used a state monad
transformer on top of that for the fresh name supply. But that's a stylistic
choice.

\speak{\studname{}} You seem to be making a big deal out of this. I
just did what I thought was the most straightforward thing to do.

@@

\subsection{Related work}

@@

\speak{\docname{}} Others have solved this problem before but I have
never seen a solution as simple as yours and was quite surprised that
it works. For example, in this paper \direct{\docname{} shows
  \studname{} the paper \cite{Generic}}, the authors use a
continuation monad to be able to reify the monadic constructs of their
language. If they were to implement your robot language they would use
the same \texttt{Program} data type as you do. However, they would not
expose that to the user of the language. Instead they would create a
type like this:

@@

\begin{small}
\begin{verbatim}
data P a =
  P (forall r. ((a -> Program r) -> Program r))
\end{verbatim}
\end{small}

\begin{dialogue}
\speak{\docname{}} The type \texttt{P} is a continuation monad so the
monad instance comes naturally. Operations on \texttt{P} are defined
using \texttt{Bind} like this:
\end{dialogue}

\begin{small}
\begin{verbatim}
while :: P BoolE -> P () -> P ()
while c b =
  P (\k -> While (runP c) (runP b) `Bind` k)
\end{verbatim}
\end{small}
%\pagebreak
\begin{small}
\begin{verbatim}
cond :: BoolE -> P () -> P () -> P ()
cond b t e =
  P (\k -> Cond b (runP t) (runP e) `Bind` k)

runP :: P a -> Program a
runP (P f) = f (\a -> Return a)
\end{verbatim}
\end{small}

\begin{dialogue}
\speak{\studname{}} I see. Continuations are quite magical to me. I
could never have come up with that technique. How do they deal with
transforming higher-order programs to first-order programs?

\speak{\docname{}} That's a good question. The transformation to first
order programs is dealt with by a library called Syntactic. It is
described in a separate paper \direct{\docname{} pulls out the paper
  \cite{axelsson2012generic} from a pile of papers}. It's hard to do
an apples to apples comparison between Syntactic and your
technique. Syntactic solves a much bigger problem than what you're
doing so it is naturally more complicated.

\speak{\studname{}} Ok.

\speak{\docname{}} There are also the papers
\cite{Farmer,sculthorpe2013constrained} which solves
the problem in a manner that is more similar to yours. They also have explicit
constructors {\tt Bind} and {\tt Return} but give them a slightly different
type which guarantees that all applications of {\tt Bind} are {\em normalised},
so that all {\tt Bind}s are right-associated. Unexpectedly, this
allows them to make instances of the {\tt Monad} class but at the same time
constrain the arguments of {\tt Bind}.
Their technique is more complicated than yours, but just as with the
case of Syntactic, they solve a more general problem.

However, there is one particular thing I like about your technique.

\speak{\studname{}} What is that?

\speak{\docname{}} Your technique is compositional.

\subsection{Compositionality}

\begin{dialogue} 
\speak{\studname{}} What does it mean that my technique is compositional?

\speak{\docname{}} The compile function is compositional because there
is one case for each constructor and each case deals with exactly one
constructor. In particular, the constructors {\tt Return} and {\tt
  Bind} are handled completely separately from all other language
constructs.

\speak{\studname{}} And that is good ?

\speak{\docname{}} Absolutely! Compositional definitions are nice
because they imply that there is no weird semantical interaction
between the constructs as they are defined independently. But it also
means that it should be possible to factor out the constructors
\texttt{Return} and \texttt{Bind} into a data type of their own. Then,
it could be combined with other data types using techniques like Data
Types \'{a} la Carte \cite{Wouter} or CompData \cite{CompData}. This
way, a language can be designed piece by piece.You can then select the
set of pieces required for a particular task or that suits a
particular brand of robots.

\speak{\studname{}} Oh! Interesting.
\end{dialogue}

\subsection{The Monad laws}

\begin{dialogue}
\speak{\docname{}} There is still one problem with your method though.

\speak{\studname{}} What's that?

\speak{\docname{}} When we make instances of the monad class we expect
certain laws to hold. \direct{\docname{} writes the laws on his whiteboard}

\begin{small}
\begin{verbatim}
m >>= return    = m
return a >>= f  = f a
(m >>= f) >>= g = m >>= \a -> f a >>= g
\end{verbatim}
\end{small}

\begin{dialogue}
\speak{\docname{}} These laws clearly don't hold for your instance. Take the
first law for instance. The left hand side will have extra \texttt{Bind} and
\texttt{Return} constructors compared to the right hand side.

\speak{\studname{}} Hmmm. I hadn't really thought about that. But adding an
extra return at the end of a computation shouldn't make a difference
in my implementation.

\speak{\docname{}} So, you're saying that your implementation actually obeys
the first monad law?

\speak{\studname{}} Well, at least when I run my programs I will never see
any difference between the left hand side and the right hand side.

\speak{\docname{}} Aha, so what you're saying is that if we compare the
semantics of programs rather than comparing the programs themselves
then we get some useful laws. \direct{\docname{} scribbles some new laws
  on the whiteboard}

\end{dialogue}
%\end{minipage}

\begin{small}
\begin{verbatim}
eval (m >>= return)    = eval m
eval (return a >>= f)  = eval (f a)
eval ((m >>= f) >>= g) = eval (m >>= \a -> f a >>= g)
\end{verbatim}
\end{small}

%\begin{minipage}[]{\linewidth}
\begin{dialogue}
\speak{\docname{}} These laws are morally the same as the monad
laws, especially if we don't let the user of the robot language ever
compare terms in the language.

\speak{\studname{}} Yes, that captures my intuition very well.

\speak{\docname{}} Ok, good. Can you prove these equations?

\speak{\studname{}} No, I don't have any experience proving programs correct.

\speak{\docname{}} Well, it shouldn't be that difficult. Let me see
what I can come up with.

\direct{\docname{} scribbles frenetically on a piece of paper for a
  couple of minutes.}

Aha! Your technique is quite general. It can reify
monads into any kind of structure which is a monoid. Return translates
into the monoid unit and bind translates into the monoid operation.

For example, in your compilation function it is important that the semantics of
\texttt{PSkip} is the identity of the semantics of \texttt{PSeq} and that
\texttt{PSeq}
is associative.
\speak{\studname{}} Ok, that sounds good. I take that as meaning the method
is quite general?

\speak{\docname{}} Yes, requiring a monoid is a very mild restriction.

Look at the time! This was interesting, I got quite carried away. We must 
round off but please get back to me if you make any more progress on your 
robot language.

\speak{\studname{}} Thanks very much for your time. Bye!
\speak{\docname{}} Thank you.

\section{Composing reifiable monadic languages} 
\label{sec:composing}

\emph{At \studname{}'s computer} \newline

\studname{} was really intrigued by the idea of compositionally
building embedded languages and after having read the papers
\docname{} showed him he decides to try his own approach to the problem.

Inspired by CompData he starts out by adding the code in figure
\ref{fig:compdata} to his file. The data type \texttt{:+:} is used for
composing languages, and the \texttt{:<:} typeclass provides coercions
so that the programmer doesn't have to worry about using the right
sequence of the \texttt{InjL} and \texttt{InjR} constructors to inject
terms into the composed language.

\begin{figure}
\begin{small}
\begin{verbatim}
data (e1 :+: e2) x a 
  = InjL (e1 x a) | InjR (e2 x a)
infixr :+:

class sub :<: sup where
  inj :: sub x a  -> sup x a

instance f :<: f where
  inj = id

instance (f :<: (f :+: g)) where
   inj = InjL

instance (f :<: h) => (f :<: (g :+: h)) where
  inj = InjR . inj 
\end{verbatim}
\end{small}
\caption{Data types from CompData for composing languages.}
\label{fig:compdata}
\end{figure}

\studname{} then starts to add data types for the different language
constructs in his robotic language. The different operations for the
robot are pleasingly easy to add.
%\pagebreak
\begin{small}
\begin{verbatim}
data MoveOp x a where 
  Move :: MoveOp x ()

data TurnOp x a where 
  TurnLeft  :: TurnOp x () 
  TurnRight :: TurnOp x ()

data SensorOp x a where 
  Sensor :: SensorOp x BoolE

data CondOp x a where 
  Cond :: x BoolE -> x () -> x () -> CondOp x ()

data WhileOp x a where 
  While :: x BoolE -> x () -> WhileOp x () 
\end{verbatim}
\end{small}

In order to test out these definitions, \studname{} next turns to
implementing a compiler. He realises that the compiler now needs to be
implemented as a class with one instance per compilable sub-language.

\begin{small}
\begin{verbatim}
runCompile :: Compile f => f a -> Prg
runCompile prg = snd $ compile s prg
  where s = newNameSupply

class Compile f where
  compile :: NameSupply -> f a -> (a, Prg)

instance Compile (MoveOp x) where
  compile _ Move = ((), PMove)

instance Compile (TurnOp x) where
  compile _ TurnLeft = ((), PTurnLeft)
  compile _ TurnRight = ((), PTurnRight)

>>>>

\section{Introduction}
\studname{} is a computer science student working in a project involving programming 
robots in a low level imperative language. However, \studname{} has a budding 
interest in functional programming using Haskell and has read ``The Haskell 
school of expression'' \cite{HUDAK}. Naturally he gets the idea to implement 
a language for robot control embedded in Haskell. \studname{} realise that the 
capabilities of their robot hardware is very similar to those of the 
robots that Hudak evaluates graphically in a grid world. There is however 
a very important difference; the embedded language needs to be compiled into 
some form that is understood by the robot. 
%\studname{} in a computer science student with a budding interest in functional 
%programming using Haskell. \studname{} is developing  
%an embedded domain specific language (EDSL) for robot control. Inspired 
%by the book ``the Haskell school of expression'' \cite{HUDAK} but instead of 
%evaluating his robot EDSL, \studname{} wants to compile the language to a form 
%he can run on actual robot hardware. 
%with 
%different motivations, \studname{} wants compile his EDSL for actual physical robots, 
%he sets out.

Guided by the capabilities of the target robot, \studname{} designs an API for 
robot programming. Amongst the capabilities of the robot are operations 
such as {\em move} that steps the robot forward and {\em turn} left and right. 
The robot also has a forward facing {\tt sensor} with which it can query the 
world and also the capabilities to execute program loops and conditionals.

The API that \studname{} designs is simple:

%\begin{figure} 
\begin{small}
\begin{verbatim} 
move :: Program () 
turnLeft :: Program () 
turnRight :: Program ()
sensor :: Program Bool 
cond :: Bool -> Program () -> Program () -> Program () 
while :: Program Bool -> Program () -> Program () 
\end{verbatim}
\end{small} 
%\label{fig:interface} 
%\caption{Proposed set of basic robot operations}
%\end{figure}

@@

%\begin{minipage}[]{\linewidth}
\vspace{5mm}

\fbox{%
\parbox{0.8\linewidth}{What \studname{} does not know at this time is that he has stumbled over the 
problem of {\em Monad reification}. Monad reification is the observing of the shape of the computation and this is necessary in order to compile to some low level representation.}
}
%\end{minipage}

\subsection{Example programs}
After listing the desired basic operations of the language, \studname{} tries to write 
a few very simple example programs.
%experiments 
%with what some of these robot controlling programs could look like once his 
%EDSL is finished.

The first program written is one called sMove, that implements a safe move 
operation. This operation can be executed by the robot even though it is facing 
an obstacle without risk of harming robot or obstacle. 
%.It feels like a good idea to \studname{} to check the sensor before trying to 
%move the robot forward. Otherwise something in either the world or the 
%robot might suffer damage.

\begin{small} 
\begin{verbatim}
sMove :: Program () 
sMove = do 
  s <- sensor 
  cond s turnRight move 
\end{verbatim}
\end{small}

The second program he writes moves the robot forward until it stands directly 
in front of an obstacle such as a wall.

@@

Since an abstract syntax representation of the robot programs is desired, \studname{} 
realises that the booleans of some of the operations needs to be replaced by 
boolean typed expressions.

\begin{small} 
\begin{verbatim}
data BoolE = Lit Bool
           | Var String
           | (:||:) BoolE BoolE
           | (:&&:)  BoolE BoolE
           | Not BoolE  
\end{verbatim} 
\end{small}

following this, the  operations that should go into the {\tt Program} data type 
feels straightforward.

\begin{small} 
\begin{verbatim}
data Program a where 
  Move :: Program () 
  TurnLeft :: Program () 
  TurnRight :: Program () 
  Sensor :: Program BoolE
  Cond :: BoolE -> Program () -> Program () 
  While :: Program BoolE -> Program () -> Program () 
\end{verbatim} 
\end{small}

\studname{} continues by naively adding constructors for {\tt Return} and 
{\tt Bind} to the {\tt Program a} data type.

@@

\section{Problem statement} 
\emph{Meanwhile in Professor \docname{}'s Office}\newline \newline

@@

\noindent\colorbox{light-gray}{
\begin{minipage}[]{0.9\linewidth}
\noindent 
Dear Professor \docname{}
\newline \newline
\noindent I am CS student working in a robot control project. Usually, we 
program our robots in C but I have developed an interest in Haskell 
programming and thought it'd be natural to try implementing an EDSL. 
I know you teach an FP course and thought id ask for your input. 
Atached to this mail is a file containing an outline of the data types I 
want to use. Does this look sensible to you? \newline \newline

@@

Prof. \docname{} opens the attachment and takes a look at the data type. He 
particularly notices the constructors {\tt Return} and {\tt Bind} and the 
{\tt Monad} instance. He shakes his head at \studname{}'s naivety and writes an 
email back:

\noindent\colorbox{light-gray}{
\begin{minipage}[]{0.9\linewidth}
\noindent 
Hello \studname{}
\newline \newline
\noindent I'm afraid your implementation of the monadic primitives can never 
work. The constructor {\tt Return} can take any arbitrary value which has 
nothing to do with the language you're designing. These values may be 
strings, binary trees of higher order functions from zygomorphisms to 
histomorphisms. The same problem goes for {\tt Bind}, it has to be able 
to handle arbitrary values. It cannot possibly work! This is a very 
difficult problem and such a naive solution is bound to fail.

\noindent For hints towards a solution to this I suggest that you follow 
this \underline{link} \cite{GillBlog}. Another idea is that you refer 
to the article ``Generic monadic constructs for embedded languages'' 
\cite{Generic}. \newline \newline

@@

\section{Implementation continues}
\emph{Back to \studname{}'s implementation efforts} \newline \newline

@@

When \docname{}'s email does arrive, \studname{} is confused. He feels he has already managed 
to compile the {\tt Program a} data type into a representation closer to that 
which is executed by the robot hardware.

@@

\noindent\colorbox{light-gray}{
\begin{minipage}[]{0.9\linewidth}
\noindent 
Hello Prof. \docname{} 
\newline \newline
\noindent Thanks for your feedback. But I think it does work! attached 
to this email (Compiler.hs) you find my attempt at compiling the 
{\tt Program a} into a representation closer to that which is executed 
by our robots. \newline \newline

@@

\section{Compilation of the monadic robot EDSL} 
\emph{\docname{} receives \studname{}'s latest email}\newline \newline

\noindent \docname{} looks through the {\tt Compiler} module and finds a data type describing
a first order representation of the robot language.

\begin{small} 
\begin{verbatim}
data Prg = PMove          
         | PTurnRight
         | PTurnLeft
         | PSensor Var    
         | PCond BoolE Prg Prg
         | PWhile Var Prg Prg 
         | PSeq Prg Prg
         | PSkip       
         | PAssign Var BoolE
\end{verbatim}
\end{small}

From the {\tt Prg} data type, \docname{} presumes that the {\tt PSeq} constructor 
will replace occurences of {\tt Bind} and that the variable binding of the 
{\tt Sensor} construct is reified using {\tt PAssign}. The conclusion is
that the representation looks sensible but he is very interested in seeing 
the {\tt Program a -> Prg} transformation.

\begin{small} 
\begin{verbatim}
compile :: Program a -> Prg
compile prg = snd $ compile' s prg 
  where 
    s = unsafePerformIO $ newEnumSupply

compile' :: Supply Int -> Program a -> (a, Prg) 
compile' s Move = ((),PMove)
compile' s TurnRight = ((),PTurnRight)
compile' s TurnLeft  = ((),PTurnLeft)
compile' s Sensor    = (Var nom,PSensor nom)
  where
    v = supplyValue s
    nom = "v" ++ show v 
compile' s (Cond b p1 p2) = ((), PCond b p1' p2') 
  where
    (s1,s2) = split2 s
    (a1,p1') = compile' s1 p1
    (a2,p2') = compile' s2 p2 
compile' s (While wp prg) = ((),PWhile nom nwp prg') 
  where
    (s1,s2,s3) = split3 s
    (b,wp') = compile' s1 wp
    (c,prg') = compile' s2 prg 
    nom  = "v" ++ (show $ supplyValue s3)
    nwp = (wp' `PSeq` PAssign nom b) 
compile' s (Return a) = (a,PSkip)
compile' s (Bind pa f) = (b, prg1 `PSeq` prg2) 
  where
    (s1,s2) = split2 s
    (a,prg1) = compile' s1 pa
    (b,prg2) = compile' s2 (f a) 
\end{verbatim}
\end{small}

\docname{} studies the code in disbelief but saves the module to his harddrive 
and tries it out on some examples. The code does indeed seem to work and 
\docname{}'s disbelief is replaced with enthusiasm, this appears to be a simple 
and conveniant way to reify monads. what's more it does so in a 
compositional way. The compilation of the {\tt Return} and {\tt Bind} cases 
are completly independent of any other construct of the language.

@@

\section{Related Work} 
\emph{Meeting in \docname{}'s office} \newline

@@

\speak{\docname{}} So, about this robot language! you gave me a bit of a 
surprise there. I was entirely sure that what you did was impossible.

@@

\speak{\docname{}} The problem is \direct{\docname{} approaches his whiteboard}
 when you try to reify the bind constructor of your representation..
\end{dialogue}

@@

\begin{dialogue}
\speak{\docname{}} you need to come up with an element of type a to pass to the 
function.

@@

\speak{\docname{}} But it is. I have never seen a solution as naive as yours 
though and was quite surprised that it works.

\speak{\docname{}} For example, in this paper \cite{Generic}, the authors use 
a continuation monad to be able to reify the monadic constructs of their 
language

\end{dialogue}
\emph{Have \docname{} outline the continuation monad approach on the whiteboard}

%\emph{Maybe here kurt can show related work and explain why these related
%approaches exist\cite{GillBlog}
%\cite{Farmer}
%\cite{Generic}
%\cite{Wouter}
%\cite{CompData}
%\cite{Compile} 
%\cite{MonadReader15}}

\begin{dialogue}
\speak{\docname{}} Your solution is that you are careful about the return types 
of other operations in your language. They are either of unit type or some 
in itself reifyable expression type. \direct{\docname{} scribbles on  his whiteboard}

@@

\emph{lots more text here}

\speak{\docname{}} About compositionality, I think that we can make this explicit 
by using an approach based on \cite{Wouter} or \cite{CompData}.

\speak{\docname{}}  Bye!  
\end{dialogue}
%\end{minipage}

\section{Execution of the compiled representation} 
\emph{The discussion turns to practical}\newline \newline 
\FloatBarrier

\speak{\docname{}} Let us see if we can implement some more interesting 
example programs in your language. Can you do spiral ?

\speak{\studname{}} No. Infinite programs are ruled out by my compilation 
function.

\speak{\docname{}} It is true that your compilation function rules out 
infinite programs but that is necessary to rule out such programs anyway 
if you want them to compile for our robot hardware. 
\speak{\docname{}} Write a spiralIn program with a well defined termination 
condition.

\direct{\studname{} opens his laptop and starts hacking..} 
\end {dialogue} 
% \end{minipage}

\begin{small}
\begin{verbatim}
spiralIn :: Int -> Program () 
spiralIn 0 = return () 
spiralIn n = do
  replicateM_ 2 $ do
    replicateM_ n move
    turnLeft
  spiralIn (n-1) 
\end{verbatim}
\end{small}

% \begin{minipage}[]{\linewidth}
\begin{dialogue}

\speak{\docname{}} yes. That should work. Do we have any way to verify this 
program.

\speak{\studname{}} We can run it through the graphical simulator!

\speak{\docname{}} Show me!

@@

\includegraphics[width=.49\linewidth]{./spiral1}
\includegraphics[width=.49\linewidth]{./spiral2}

@@

\speak{\docname{}} Nice!

\speak{\docname{}} Now, let us try something more advanced. Can you program 
the robot to always keep a wall on his left hand side?

@@

\begin{small}
\begin{verbatim} 
followWall :: Program () 
followWall = do
  while (return true) $ do
    b <- checkLeft
    cond b sMove (turnLeft >> move) 
\end{verbatim}
\end{small}

\begin{dialogue} 
\speak{\docname{}} Yes. that looks about right. Let us see it. 
\end{dialogue}

\includegraphics[width=.49\linewidth]{./wall1}
\includegraphics[width=.49\linewidth]{./wall2}
